### 2021-05-24 chaps/thesis/chap1/sections/article.tex
✏️ fix typo in citation and remove red commentary

@@

For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $y_{\max} = \min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t)$. As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that $i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D}$, where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. 
Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability.
Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{Crouseilles:2018} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. \textcolor{black}{In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.}
\end{itemize}

@@

{\color{red}
Finally, we give an illustration of the above comments by running \eqref{vp_linear} ($d=b=1$ and $v_{\max}=3$) 
with a discontinuous initial data to study the impact of high-frequency on the stability of exponential schemes coupled 
with a centered scheme in $v$. 
The initial condition is 
$$
f_0(x, v) = 1 \mbox{ if } \sqrt{(x-\pi)^2+v^2} \leq 1, \mbox{ and } 0  \mbox{ elsewhere}. 
$$
From the stability analysis we know that for each $k$ it holds that $\bar{f}^{n+1}_{k,m} = \phi(z)\bar{f}^{n}_{k,m}$,
where $\phi$ is the stability function. 
Our study enables us to estimate the amplification factor $|\phi(z)|$ by $(1+\varepsilon)$ (uniformly with respect to $k$)  
so that $|\bar{f}^{n+1}_{k,m}|^2 \leq  (1+\varepsilon)^2 |\bar{f}^{n}_{k,m}|^2$. Hence, for a given $\varepsilon$, 
we consider a time step according to Table \ref{tab:ymax_expo} and run 
two exponential schemes, namely Hochbruck-Ostermann and Cox-Matthews,   
for $100$ timesteps. After each time step
we compute $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$. 
The results can be found in Figure \ref{instab}.   
\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[scale=0.5]{\localPath/figures/error_growup_HO.pdf} & 
\includegraphics[scale=0.5]{\localPath/figures/error_growup_CM.pdf} 
\end{tabular}
\caption{Evolution of $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$ and of $(1+\varepsilon)^{2n}$ as a function of $n$ for different values of 
$\varepsilon$. Left: Hochbruck-Ostermann scheme. Right: Cox-Matthews scheme. }
\label{instab}
\end{figure}
First, we have a numerical confirmation that $(1+\varepsilon)$ is an estimate of the amplification factor 
(the ratio $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$  always lies under the curve $n\mapsto (1+\varepsilon)^{2n}$).  
Second, since the number of time steps is fixed, for $\varepsilon=0.1$ the simulation becomes unstable for $n\geq 20$. 
As soon as $\varepsilon$ is small enough, the simulation is stable. 
%As mentioned above, if $\varepsilon$ as $\varepsilon =C \Delta t = Ct_{final}/n$  
%(with $\Delta t=y_{\max} \Delta v$, $y_{\max}$ being given by Table \ref{tab:ymax_expo} and $\Delta v=6/81$), 
%the scheme is  stable and considering a final time $t_{final} = n\Delta t$, the linear analysis leads to 
%\begin{eqnarray*}
%\|{f}^{n}\|_{\ell^2} &=& \|\bar{f}^{n}\|_{\ell^2} \leq  (1+\varepsilon)^n \|\bar{f}^{0}\|_{\ell^2} =  (1+C\Delta t)^{t_{final}/\Delta t} \|\bar{f}^{0}\|_{\ell^2} \nonumber\\
%&\leq & \exp(Ct_{final}) \|\bar{f}^{0}\|_{\ell^2} = \exp(Ct_{final}) \|{f}^{0}\|_{\ell^2}, 
%\end{eqnarray*}
%which gives some practical informations on the stability of the methods since $C$ and $t_{final}$ are known a priori. 
}

@@

In Figure \ref{fig:mass_energy}, the time history of the relative error of the total mass and of the total energy 
are displayed with the phase space discretization $64\times 64\times 64\times 128$ and 
using Lawson($RK(4,4)$) and Cox--Matthews time integrators (the perturbation formulation is used here). Since mass is a linear invariant it is preserved, up to machine precision, by the exponential integrator, see \cite{le2015KP}. We also observe good conservation of energy even in the nonlinear phase, which confirms the excellent behavior of the methods.

>>>>

% !TEX root = ../../main.tex

@@

For the exponential integrators the domain of stability is very sensitive to the value of $(a\Delta t)$. To get an idea of what we can expect, we consider the quantity $y_{\max}=\min_{(a\Delta t)\in\mathbb{R}} \; y^{exp}_{\max}(a \Delta t)$. As before, $y^{exp}_{\max}(a\Delta t)$ is the largest value such that : $i(-y^{exp}_{\max}(a\Delta t), y^{exp}_{\max}(a\Delta t))\subset {\cal D}$, where ${\cal D}$ is the domain of stability for the chosen exponential time integrator for a given $(a\Delta t)$. 
Even for relatively simple numerical methods it is not possible to compute this quantity analytically. Thus, we resort to numerical approximations. Unfortunately, it turns out that for most exponential integrators this value is zero. This can be appreciated by considering Figure \ref{fig:expRK_sd} once more. Clearly, there are values of $(a \Delta t)$ such that no relevant part of the imaginary axis (or only half the imaginary axis) is part of the domain of stability.
Thus, most exponential integrators are unstable in the von Neumann sense. However, this is not what we observe in practice. In fact, already the results presented in \cite{Crouseilles:2018} indicate that we can successfully run numerical simulations using, for example, the Cox--Matthews scheme. There are two major points to consider here
\begin{itemize}
    \item The $y_{\max}$ obtained is a worst case estimate. In fact, we know that for $\Delta t \to 0$ we regain the stability of the underlying Runge--Kutta method. Thus, for small $(a \Delta t)$ the methods is expected to work well.
    \item As is usually done we have mandated that $\vert \phi(z) \vert \leq 1$. However, strictly speaking this is not necessary for practical simulation. If we assume that $\vert \phi(z) \vert \leq 1+\varepsilon$ and we take $n$ steps the amplification of the error is given by $(1+\varepsilon)^n$. In the limit $n \to +\infty$ this quantity diverges. However, since we usually do not take infinitely small time steps we still can hope to obtain a relatively accurate approximation, especially if $\varepsilon$ is small. \textcolor{black}{In particular, if $\varepsilon = C \Delta t = C t_{final}/n$ (where $t_{final}$ denotes the final time and $n$ the number of iterations) we have $(1+\varepsilon)^n = (1+\tfrac{C t_{final}}{n} )^n \leq \exp(C t_{final})$ and thus the scheme is stable, while the error constant is increased by $\exp(C t_{final})$.}
\end{itemize}

@@

%{\color{red}
Finally, we give an illustration of the above comments by running \eqref{vp_linear} ($d=b=1$ and $v_{\max}=3$) 
with a discontinuous initial data to study the impact of high-frequency on the stability of exponential schemes coupled 
with a centered scheme in $v$. 
The initial condition is 
$$
f_0(x, v) = 1 \mbox{ if } \sqrt{(x-\pi)^2+v^2} \leq 1, \mbox{ and } 0  \mbox{ elsewhere}. 
$$
From the stability analysis we know that for each $k$ it holds that $\bar{f}^{n+1}_{k,m} = \phi(z)\bar{f}^{n}_{k,m}$,
where $\phi$ is the stability function. 
Our study enables us to estimate the amplification factor $|\phi(z)|$ by $(1+\varepsilon)$ (uniformly with respect to $k$)  
so that $|\bar{f}^{n+1}_{k,m}|^2 \leq  (1+\varepsilon)^2 |\bar{f}^{n}_{k,m}|^2$. Hence, for a given $\varepsilon$, 
we consider a time step according to Table \ref{tab:ymax_expo} and run 
two exponential schemes, namely Hochbruck-Ostermann and Cox-Matthews,   
for $100$ timesteps. After each time step
we compute $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$. 
The results can be found in Figure \ref{instab}.   
\begin{figure}
\centering
\begin{tabular}{cc}
\includegraphics[scale=0.5]{\localPath/figures/error_growup_HO.pdf} & 
\includegraphics[scale=0.5]{\localPath/figures/error_growup_CM.pdf} 
\end{tabular}
\caption{Evolution of $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$ and of $(1+\varepsilon)^{2n}$ as a function of $n$ for different values of 
$\varepsilon$. Left: Hochbruck-Ostermann scheme. Right: Cox-Matthews scheme. }
\label{instab}
\end{figure}
First, we have a numerical confirmation that $(1+\varepsilon)$ is an estimate of the amplification factor 
(the ratio $\| f^n\|^2_{\ell^2} / \| f^0\|^2_{\ell^2}$  always lies under the curve $n\mapsto (1+\varepsilon)^{2n}$).  
Second, since the number of time steps is fixed, for $\varepsilon=0.1$ the simulation becomes unstable for $n\geq 20$. 
As soon as $\varepsilon$ is small enough, the simulation is stable. 
%As mentioned above, if $\varepsilon$ as $\varepsilon =C \Delta t = Ct_{final}/n$  
%(with $\Delta t=y_{\max} \Delta v$, $y_{\max}$ being given by Table \ref{tab:ymax_expo} and $\Delta v=6/81$), 
%the scheme is  stable and considering a final time $t_{final} = n\Delta t$, the linear analysis leads to 
%\begin{eqnarray*}
%\|{f}^{n}\|_{\ell^2} &=& \|\bar{f}^{n}\|_{\ell^2} \leq  (1+\varepsilon)^n \|\bar{f}^{0}\|_{\ell^2} =  (1+C\Delta t)^{t_{final}/\Delta t} \|\bar{f}^{0}\|_{\ell^2} \nonumber\\
%&\leq & \exp(Ct_{final}) \|\bar{f}^{0}\|_{\ell^2} = \exp(Ct_{final}) \|{f}^{0}\|_{\ell^2}, 
%\end{eqnarray*}
%which gives some practical informations on the stability of the methods since $C$ and $t_{final}$ are known a priori. 
%}

@@

In Figure \ref{fig:mass_energy}, the time history of the relative error of the total mass and of the total energy 
are displayed with the phase space discretization $64\times 64\times 64\times 128$ and 
using Lawson($RK(4,4)$) and Cox--Matthews time integrators (the perturbation formulation is used here). Since mass is a linear invariant it is preserved, up to machine precision, by the exponential integrator, see \cite{Einkemmer:2015}. We also observe good conservation of energy even in the nonlinear phase, which confirms the excellent behavior of the methods.
